Do we need detailed demographic data to forecast plant population responses to climate change?

### 2015-06-18 manuscript/manuscript.tex
updates ms discussion

\documentclass[12pt,]{article}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
% Redefine \includegraphics so that, unless explicit options are
% given, the image width will not exceed the width of the page.
% Images get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\ScaleIfNeeded{%
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother
\let\Oldincludegraphics\includegraphics
{%
 \catcode`\@=11\relax%
 \gdef\includegraphics{\@ifnextchar[{\Oldincludegraphics}{\Oldincludegraphics[width=\ScaleIfNeeded]}}%
}%
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

\author{}
\date{}
\usepackage{lineno}
\linenumbers
\usepackage{setspace}
\doublespacing

@@

\normalsize

\section{Do we need detailed demographic data to forecast the impacts of
climate change on
populations?}\label{do-we-need-detailed-demographic-data-to-forecast-the-impacts-of-climate-change-on-populations}

\subsubsection[Andrew T. Tredennick and Peter B. Adler]{Andrew T.
Tredennick\footnote{Corresponding author:
  \href{mailto:atredenn@gmail.com}{\href{mailto:atredenn@gmail.com}{atredenn@gmail.com}}}
and Peter B.
Adler}\label{andrew-t.-tredennickcorrauth-and-peter-b.-adler}

@@

Our demographic data comes from the Fort Keogh Livestock and Range
Research Laboratory in eastern Montana's northern mixed prairie near
Miles City, Montana, USA ($46^{\circ}$ 19' N, $105^{\circ}$ 48' W). The
dataset is freely available on Ecological Archives\footnote{\url{http://esapubs.org/archive/ecol/E092/143/}}
(Anderson et al. 2011) , and interested readers should refer to the
metadata therein for a complete description. The site is about 800 m
above sea level and mean annual precipitation (1878-2009) is 334 mm,
with most annual precipitation falling from April through September. The
site is grass dominated and, for the purposes of our study, we focus on
the four most abundant graminoid species: \emph{Bouteloua gracilis}
(BOGR), \emph{Hesperostipa comata} (HECO), \emph{Pascopyrum smithii}
(PASM), and \emph{Poa secunda} (POSE).

From 1932 to 1945 individual plants were identified and mapped annually
in 44 1-$\text{m}^2$ quadrats using a pantograph. The quadrats were
distributed in six pastures, each assigned a grazing treatment of light
(1.24 ha/animal unit month), moderate (0.92 ha/aum), and heavy (0.76
ha/aum) stocking rates (two pastures per treatment). In this analysis we
account for potential differences among the grazing treatments, but do
not focus on grazing$\times$climate interactions. The annual maps of the
quadrats were digitized and the fates of individual plants tracked and
extracted using a computer program. Daily climate data, which we
aggregated into climate variables of interest, are available for the
duration of the data collection period (1932 - 1945) from the Miles City
airport, Wiley Field, 9 km from the study site.

In this paper, we model populations based on two levels of data:
individual and quadrat (Fig. 1). The individual data is the ``raw''
data. For the quadrat level we data we simply sum individual areal cover
for each quadrat by species. This is equivalent to a perfect census of
quadrat percent cover, so we do not need to consider measurement error.
Based on these two datasets we can compare population models built using
individual level data and aggregated quadrat level data.

@@

Equilibrium cover from both models was sensitive to climate (Fig. 2a-d).
The IPM projected percent changes in equilibrium cover from -3 - 8\% for
\emph{B. gracilis}, -4 - 3\% for \emph{H. comata}, -15 - 9\% for
\emph{P. smithii}, and -17 - 53\% for \emph{P. secunda}. The QBM
projected opposite and greater percent changes in equilibrium cover for
\emph{B. gracilis} (-63 - 30\%) and \emph{H. comata} (-50 - -18\%; Fig.
2a-b). For \emph{P. smithii}, the QBM projected opposite changes in
equilibrium cover than the IPM, but of similar magnitude (-5 - 6\%; Fig.
2c). \emph{P. secunda} was the only species that the IPM and QBM made
projections of the same sign and somewhat similar magnitude (Fig. 2d).
As expected based on model validation (Table 1), IPM projections were
more uncertain than QBM projections for all species and all climate
change scenarios (Fig. 2e-h).

The response of a population to climate change is a result of the
aggregate effects of climate on individual vital rates. Since the IPM
approach relies on vital rate regressions, we were able to quantify the
sensitivity of each vital rate in isolation and in pairwise
combinations. Species showed similar trends (Fig. 3). Growth was the
most sensitive vital rate for all species, showing a negative response
to increased precipitation, and stronger positive response to increased
temperature, and a mostly positive response when both climate factors
are increased (Fig. 3). \emph{B. gracilis} survival rates were sensitive
to temperature, resulting in an increase in plant cover under increased
temperature (Fig. 3a). In isolation, recruitment and survival were
insensitive to climate factors for \emph{H. comata} (Fig. 3b). Survival
and recruitment of \emph{P. smithii} were both sensitive, negatively, to
temperature and precipitation (Fig. 3c). \emph{P. secunda} equilibrium
cover was sensitive to the climate effects on survival and recruitment,
showing a negative effect on both vital rates for increased precipition,
but a strong positive effect on survival with increased temperature
(Fig. 3d). The climate impact of recruitment on equilibrium cover was
negative for precipitation and temperature increases (Fig. 3d).

@@

As a potential remedy to the ``data dearth'' problem, Queenborough et
al. (2011) and Freckleton et al. (2011), building on work by Taylor and
Hastings (2004), advocate a ``density-structured'' modeling approach.
Such models do not require individual level demographic data. Indeed,
Queenborough et al. (2011) provide empirical evidence that
density-structured models adequately described population dynamics. The
results from density-structured models are not as precise as those from
traditional population models, but the loss in precision is traded off
with a gain in data. The study by Queenborough et al. (2011) included
data from 500 fields (4 hectares each) in 49 farms, all collected by two
people in 6 weeks. This is far more data from a far greater spatial
extent than possible if measuring individual plant demography (in a
world of limited time and money, at least). The appeal of
density-structured approaches is clear.

@@

This result is not as dire as it seems when we consider model
uncerainty. Yes, the QBM failed to match IPM forecasts under perturbed
climate, but the uncertainties around those forecasts were so great
(Fig. 2e-h) that we cannot say that the two model's forecasts were
different with any level of confidence. This offers hope that with
``enough'' population level data one will be able to reduce parameter
uncertainty and thus model forecasts. This remains an open question.

@@

Our intuition was that species with strong climate-dependence on vital
rates not well resolved at the aggregate level would result in different
forecasts from the two models. For example, survival is very size
dependent: smaller individuals have a higher probability of death (Chu
and Adler 2014). At the same time, a single small individual contributes
relatively little to percent cover estimates at the plot scale. So, if
survival of individuals was positively impacted by temperature
increases, for example, we would expect to detect this signal in the
individual level data but not in the aggregate level data. However, the
species with recruitment and survival rates most sensitive to climate
factors (\emph{P. secunda}; Fig. 3d) was also the only species for which
the IPM and QBM made consistent forecasts (Fig. 2d).

Individual growth was the most sensitive vital rate to climate factors
for all species (Fig. 3). Individual plant growth, if survival of large
plants is high, likely plays the biggest role in year-to-year percent
cover at the plot level. Therefore, we would naively expect forecasts
from the two models to be consistent. Since this was not the case, and
other vital rates were relatively insensitive to

@@

Our goal was not make any explicit forecast for the future state of
these populations based on predicted climate change. But our results
highlight the state of affairs in ecology when it comes to forecasting
the impacts of climate change. The analysis we conducted here could be
considered, with some exceptions of course, at the forefront of
ecological forecasting in terms of the statistical approach employed
(hierarchical Bayesian), the type of population model we used
(stochastic IPM with parameter uncertainty), and the amount of data we
had at our disposal (14 years of individual-level data). Yet, model
predictions proved so uncertain that any forecast, when bounded with
uncertainty, would be at best not useful and at worst meaningless.

Something about fitting the models\ldots{}then cite Britta's paper:
20-25 years needed!

@@

\emph{NOTES}: The IPM MAE is significantly lower at $\alpha=0.05$ for
\emph{B. gracilis} (\emph{P} = 0.0012), \emph{H. comata} (\emph{P} =
4.0586 × 10-8), and \emph{P. smithii} (\emph{P} = 3.183 × 10-5). MAEs
are statisticially similar between models for \emph{P. secunda}
(\emph{P} = 0.0922).

@@

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-figure_1.pdf}
\caption{Work flow of the data aggregation, model fitting, and
population simulating.}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-figure_2.pdf}
\caption{Proportional change in species' mean cover caused by a 1\%
increase in observed precipitation (+ppt), temperature (+temp), or both
(+ppt\&temp) as predicted by the individual-based IPM and the
aggregate-based QBM. Top panels show the mean predicted proportional
change in cover; lower panels show the 90\% quantiles of predicted
proportional cover change. Mean predictions and their associated
uncertainties are shown in separate panels for visual clarity.}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-figure_3.pdf}
\caption{Sensitivity of equilibrium cover to a 1\% increase in
precipitation (+ppt), temperature (+temp), or both (+ppt\&temp) applied
to individual and combined vital rates. For example, the points
associated with G show the median cover from IPM simulations where a
climate perturbation is applied only to the growth regression climate
covariates.}
\end{figure}

>>>>

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}
\setlength{\droptitle}{-2em}
  \title{}
  \pretitle{\vspace{\droptitle}}
  \posttitle{}
  \author{}
  \preauthor{}\postauthor{}
  \date{}
  \predate{}\postdate{}

\usepackage{lineno}
\linenumbers
\usepackage{setspace}
\doublespacing

@@

\maketitle

\section{Do we need detailed demographic data to forecast the impacts of
climate change on plant
populations?}\label{do-we-need-detailed-demographic-data-to-forecast-the-impacts-of-climate-change-on-plant-populations}

\subsubsection[Andrew T. Tredennick and Peter B. Adler]{Andrew T.
Tredennick\footnote{Corresponding author:
  \href{mailto:atredenn@gmail.com}{\href{mailto:atredenn@gmail.com}{\nolinkurl{atredenn@gmail.com}}}}
and Peter B.
Adler}\label{andrew-t.-tredennickcorrauth-and-peter-b.-adler}

@@

Our demographic data comes from the Fort Keogh Livestock and Range
Research Laboratory in eastern Montana's northern mixed prairie near
Miles City, Montana, USA (\(46^{\circ}\) 19' N, \(105^{\circ}\) 48' W).
The dataset is freely available on Ecological Archives\footnote{\url{http://esapubs.org/archive/ecol/E092/143/}}
(Anderson et al. 2011) , and interested readers should refer to the
metadata therein for a complete description. The site is about 800 m
above sea level and mean annual precipitation (1878-2009) is 334 mm,
with most annual precipitation falling from April through September. The
site is grass dominated and, for the purposes of our study, we focus on
the four most abundant graminoid species: \emph{Bouteloua gracilis}
(BOGR), \emph{Hesperostipa comata} (HECO), \emph{Pascopyrum smithii}
(PASM), and \emph{Poa secunda} (POSE) (Fig. 1).

From 1932 to 1945 individual plants were identified and mapped annually
in 44 1-\(\text{m}^2\) quadrats using a pantograph. The quadrats were
distributed in six pastures, each assigned a grazing treatment of light
(1.24 ha/animal unit month), moderate (0.92 ha/aum), and heavy (0.76
ha/aum) stocking rates (two pastures per treatment). In this analysis we
account for potential differences among the grazing treatments, but do
not focus on grazing\(\times\)climate interactions. The annual maps of
the quadrats were digitized and the fates of individual plants tracked
and extracted using a computer program. Daily climate data, which we
aggregated into climate variables of interest, are available for the
duration of the data collection period (1932 - 1945) from the Miles City
airport, Wiley Field, 9 km from the study site.

In this paper, we model populations based on two levels of data:
individual and quadrat (Fig. 2). The individual data is the ``raw''
data. For the quadrat level we data we simply sum individual areal cover
for each quadrat by species. This is equivalent to a perfect census of
quadrat percent cover, so we do not need to consider measurement error.
Based on these two datasets we can compare population models built using
individual level data and aggregated quadrat level data.

@@

Equilibrium cover from both models was sensitive to climate (Fig. 3a-d).
The IPM projected percent changes in equilibrium cover from -3 - 8\% for
\emph{B. gracilis}, -4 - 3\% for \emph{H. comata}, -15 - 9\% for
\emph{P. smithii}, and -17 - 53\% for \emph{P. secunda}. The QBM
projected opposite and greater percent changes in equilibrium cover for
\emph{B. gracilis} (-63 - 30\%) and \emph{H. comata} (-50 - -18\%; Fig.
3a-b). For \emph{P. smithii}, the QBM projected opposite changes in
equilibrium cover than the IPM, but of similar magnitude (-5 - 6\%; Fig.
3c). \emph{P. secunda} was the only species that the IPM and QBM made
projections of the same sign and somewhat similar magnitude (Fig. 3d).
As expected based on model validation (Table 1), QBM projections were
more uncertain than IPM projections for all species except \emph{P.
smithiii} (Fig. 3e-h).

The response of a population to climate change is a result of the
aggregate effects of climate on individual vital rates. Since the IPM
approach relies on vital rate regressions, we were able to quantify the
sensitivity of each vital rate in isolation and in pairwise
combinations. Species showed similar trends (Fig. 4). Growth was the
most sensitive vital rate for all species, showing a negative response
to increased precipitation, and stronger positive response to increased
temperature, and a mostly positive response when both climate factors
are increased (Fig. 4). \emph{B. gracilis} survival rates were sensitive
to temperature, resulting in an increase in plant cover under increased
temperature (Fig. 4a). In isolation, recruitment and survival were
insensitive to climate factors for \emph{H. comata} (Fig. 4b). Survival
and recruitment of \emph{P. smithii} were both sensitive, negatively, to
temperature and precipitation (Fig. 4c). \emph{P. secunda} equilibrium
cover was sensitive to the climate effects on survival and recruitment,
showing a negative effect on both vital rates for increased precipition,
but a strong positive effect on survival with increased temperature
(Fig. 4d). The climate impact of recruitment on equilibrium cover was
negative for precipitation and temperature increases (Fig. 4d).

@@

As a potential remedy to the ``data dearth'' problem, Queenborough et
al. (2011) and Freckleton et al. (2011), building on work by Taylor and
Hastings (2004), advocate a ``density-structured'' modeling approach.
Such models do not require individual level demographic data and can
adequately describe population dynamics (Queenborough et al. 2011). The
results from density-structured models are not as precise as those from
traditional population models, but the loss in precision is traded off
with a gain in data. The study by Queenborough et al. (2011) included
data from 500 fields (4 hectares each) in 49 farms, all collected by two
people in 6 weeks. This is far more data from a far greater spatial
extent than possible if measuring individual plant demography (in a
world of limited time and money, at least). The appeal of
density-structured approaches is clear.

@@

This result is not as dire as it seems when we consider model
uncerainty. Yes, the QBM failed to match IPM forecasts under perturbed
climate, but the uncertainties around those forecasts were so great
(Fig. 3e-h) that we cannot say that the two model's forecasts were
different with any level of confidence. This offers hope that with
``enough'' population level data one will be able to reduce parameter
uncertainty and thus model forecasts. This remains an open question.

@@

Our intuition was that species with strong climate-dependence on vital
rates not well resolved at the aggregate level would result in different
forecasts from the two models. For example, survival is very size
dependent: smaller individuals have a higher probability of death (Chu
and Adler 2014). At the same time, a single small individual contributes
relatively little to percent cover estimates at the plot scale. So, if
survival of individuals was positively impacted by temperature
increases, for example, we would expect to detect this signal in the
individual level data but not in the aggregate level data.
\textbf{PETER: do you think this is the right direction to go with
this?}

@@

Our goal was not make any explicit forecast for the future state of
these populations based on predicted climate change. But our results
highlight the state of affairs in ecology when it comes to forecasting
the impacts of climate change. The analysis we conducted here could be
considered, with some exceptions of course, at the forefront of
ecological forecasting in terms of the statistical approach employed
(hierarchical Bayesian), the type of population model we used
(stochastic IPM with parameter uncertainty), and the amount of data we
had at our disposal (14 years of individual-level data). Yet, model
predictions proved so uncertain that any forecast, when bounded with
uncertainty, would be at best not useful and at worst meaningless. How
might we improve on this state of affairs?

First, forecasts could be improved by matching the spatial scale of
predictor variables with the spatial scale of observations. One of the
major limitations of the models we fit here is that the climate data are
at a much larger scale than the plot and individual level observations
of plant cover and size. Climate covariates only vary by year, with no
spatial variability within years. Thus, even if we fit models to
individual level data, we are missing the key interaction point between
weather and individual plants. \textbf{PETER: any good citations here
for decent population models fit with fine-scale climate/environmental
data??}

Second, accurately detecting climate signals will take even longer time
series. Recent theoretical work on detecting climate signals in noisy
data suggests that even advanced approaches to parameter fitting like
LASSO and random forest require 20-25 year time series (Teller et al.,
in review).

@@

\emph{NOTES}: The IPM MAE is significantly lower at \(\alpha=0.05\) for
\emph{B. gracilis} (\emph{P} = 0.0012), \emph{H. comata} (\emph{P} =
4.0586 × 10-8), and \emph{P. smithii} (\emph{P} = 3.183 × 10-5). MAEs
are statisticially similar between models for \emph{P. secunda}
(\emph{P} = 0.0922).

@@

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-figure_1.pdf}
\caption{Time series of average percent cover of our four focal species:
\emph{Bouteloua gracilis} (BOGR), \emph{Hesperostipa comata} (HECO),
\emph{Pascopyrum smithii} (PASM), and \emph{Poa secunda} (POSE).}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-figure_2.pdf}
\caption{Work flow of the data aggregation, model fitting, and
population simulating.}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-figure_3.pdf}
\caption{Proportional change in species' mean cover caused by a 1\%
increase in observed precipitation (+ppt), temperature (+temp), or both
(+ppt\&temp) as predicted by the individual-based IPM and the
aggregate-based QBM. Top panels show the mean predicted proportional
change in cover; lower panels show the 90\% quantiles of predicted
proportional cover change. Mean predictions and their associated
uncertainties are shown in separate panels for visual clarity.}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-figure_4.pdf}
\caption{Sensitivity of equilibrium cover to a 1\% increase in
precipitation (+ppt), temperature (+temp), or both (+ppt\&temp) applied
to individual and combined vital rates. For example, the points
associated with G show the median cover from IPM simulations where a
climate perturbation is applied only to the growth regression climate
covariates.}
\end{figure}
